---
title: "Mass Array Error Calculations"
subtitle: "Part one of three"
author: "Bart Huntley"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

### Introduction
This rmarkdown document is the first of 3 that will process bat scat mass array data. 

The rmarkdown documents are as follows:

1. Error calculations (batscat1.Rmd) - *this document*
2. Calculating groups and majority cases (batscat2.Rmd)
3. Final visualisation and export to structure format (batscat3.Rmd)

These documents will show all code to allow for transparency. There will be sections that require input from the **USER** prior to running. It is intended that the html output accompany any csv output to act as a reproducible record.

#### Libraries
```{r, libraries, eval=TRUE, message=FALSE, warning=FALSE}
# required packages 
list.of.packages <- c("tidyverse", "janitor")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# load packages
library(tidyverse)
library(janitor)
```

#### 1 - Get the raw data
**USER** Change the filepath and name of raw data as necessary. This block ingests the data cleans names and makes ready for further analysis.
```{r, raw_data_in, eval=TRUE, message=FALSE, warning=FALSE}
# file path to raw data
fp_in <- "../source"

# name of raw data csv
n <- "MassArray2.csv"

# construct full name
csvdat <- file.path(fp_in, n)

# clean names and toss tissue samples in dataset make blanks NA
na_dat <- read_csv(csvdat, na = "Fail") %>% 
  janitor::clean_names() %>%
  dplyr::filter(sample_type == "Scat")

# same data but retain blanks (need for amplification rate)
all_dat <- read_csv(csvdat) %>% 
  janitor::clean_names() %>%
  dplyr::filter(sample_type == "Scat") %>%
  tidyr::pivot_longer(cols = starts_with("x"),
                      names_to = "marker",
                      values_to = "val") %>%
  dplyr::mutate(val = ifelse(val == "Fail", "", val)) %>%
  tidyr::pivot_wider(names_from = marker,
                     values_from = val)

# check
all_dat[1:10, 1:6]
```

#### 2 - Error calculations
This code block produces 3 important datasets:

1. `results_out` - for each sample the following errors are calculated:
    + average amplification rate
    + allele error
    + locus error
    + allelic drop out
    + false allele
2. `summaries` - mean and standard error for the above errors for whole dataset
3. `num_out` - taking into account the errors, recodes letters to numerical output
```{r, calc_errs, eval=TRUE, message=FALSE, warning=FALSE}
# just variables
df_loci <- na_dat  %>%
  dplyr::select(starts_with("X")) # keep loci columns for calcs

# clean with blanks
cl_dat_bl <- all_dat[rowSums(is.na(df_loci)) != ncol(df_loci),]

# clean with na's
cl_dat_na <- na_dat[rowSums(is.na(df_loci)) != ncol(df_loci),]

# merge alleles function for grouping step
f1 <- function(x){trimws(paste(x, collapse = ''))}

# set up for loop variables stuff to use in loop
lgth_vars <- dim(df_loci)[2]
u_samp <- unique(unlist(cl_dat_bl[, 1]))
samp_names <- paste0("s", u_samp)

# set up data frames to hold results
results_out <- data.frame(stringsAsFactors = FALSE)
num_out <- data.frame(stringsAsFactors = FALSE)

# loop and calculate interim and end results
for(i in seq_along(u_samp)){
  # using na data
  avg_amp_rate <- cl_dat_na %>%
    dplyr::filter(sample == u_samp[i]) %>%
    tidyr::pivot_longer(cols = starts_with("x"),
                        names_to = "vars",
                        values_to = "vals") %>%
    dplyr::group_by(rep) %>%
    dplyr::summarise(amp_rate = sum(!is.na(vals))/lgth_vars) %>%
    dplyr::summarise(avg_amp_rate = mean(amp_rate))
  
  # interim data (creating new variables to calculate errors)
  d <- cl_dat_bl %>%
    dplyr::filter(sample == u_samp[i]) %>%
    dplyr::group_by(sample) %>%
    dplyr::summarise_all(.funs = f1) %>%
    dplyr::ungroup() %>%
    tidyr::pivot_longer(cols = starts_with("x"),
                        names_to = "vars",
                        values_to = "vals") %>%
    dplyr::mutate(
      f = case_when(
        nchar(vals) == 4 ~ vals,
        TRUE ~ ""),
      c = nchar(f),
      loc_err = ifelse(c == 4 & substr(f, 1, 2) != substr(f, 3, 4),
                       0.5, 0),
      p1 = substr(f, 1, 1),
      p2 = substr(f, 2 ,2),
      p3 = substr(f, 3, 3),
      p4 = substr(f, 4 ,4),
      a_mm1 = ifelse(c == 4 & p1 != p3, 1, 0),
      a_mm2 = ifelse(c == 4 & p2 != p4, 1, 0),
      a_err = (a_mm1 + a_mm2)/4, 
      het_only = case_when(
        p1 == p2 & p3 == p4 ~ 0,
        TRUE ~ 1),
      a_drop = case_when(
        het_only == 1 & substr(f, 1, 2) != substr(f, 3, 4) ~ 0.5,
        TRUE ~ 0),
      fa = case_when(
        c == 4 & het_only == 0 & substr(f, 1 ,2) != substr(f, 3 ,4) ~ 0.5,
        TRUE ~ 0))
  
  # take interim data and use to convert to numerical values for export
  num_vals <- d %>%
    dplyr::mutate(new_var = case_when(
      nchar(vals) == 2 ~ substr(vals, 1, 2),
      nchar(vals) == 4 & substr(vals, 1, 2) == substr(vals, 3, 4) ~ substr(vals, 1, 2),
      TRUE ~ "NA"
    )) %>%
    dplyr::mutate(to_recode = case_when(
      new_var == "NA" ~ 0,
      TRUE ~ 1
    ),
    a1 = ifelse(to_recode == 1, substr(vals, 1, 1), NA),
    b1 = ifelse(to_recode == 1, substr(vals, 2, 2), NA),
    a = case_when(
      a1 == "A" ~ 110,
      a1 == "T" ~ 120,
      a1 == "G" ~ 130,
      a1 == "C" ~ 140
    ),
    b = case_when(
      b1 == "A" ~ 110,
      b1 == "T" ~ 120,
      b1 == "G" ~ 130,
      b1 == "C" ~ 140)
    ) %>%
    dplyr::select(sample, vars, a, b) %>%
    tidyr::pivot_longer(cols = c("a", "b"),
                        names_to = "locus",
                        values_to = "value") %>%
    dplyr::arrange(vars, locus) %>%
    tidyr::pivot_wider(names_from = c(vars, locus), values_from = value)
  
  # take interim data and calculate errors for export
  results <- d %>%
    dplyr::summarise(locus_error = sum(loc_err)/sum(c == 4),
                     allele_error = sum(a_err)/sum(c == 4),
                     allelic_drop_out = sum(a_drop)/sum(het_only),
                     false_allele = sum(fa)/(sum(c)/4)) %>%
    dplyr::bind_cols(avg_amp_rate) %>%
    dplyr::mutate(sample = samp_names[i]) %>%
    dplyr::select(sample, avg_amp_rate, allele_error, locus_error, allelic_drop_out, 
                  false_allele)
  
  # bind exports to appropriate result data frames
  results_out <- dplyr::bind_rows(results_out, results)
  num_out <- dplyr::bind_rows(num_out, num_vals)
}

# take the error results and further summarise
summaries <- results_out %>%
  tidyr::pivot_longer(cols = -sample,
                      names_to = "error",
                      values_to = "value") %>%
  dplyr::group_by(error) %>%
  dplyr::summarise(avg = mean(value, na.rm = TRUE),
                   se = sd(value, na.rm = TRUE)/sqrt(n()))

# check
summaries
```

#### 3 - Export results to csv
This code block will write the datasets from above to csv. The `num_out` dataset will be required for continued analysis with the remaining rmarkdown documents.

**USER** outputs will be written to `/results`. Ensure you adjust the filepath to suit your situation.
```{r, export, eval=TRUE, message=FALSE, warning=FALSE}
# desired file path for output
fp_out <- "."

# create results directory if doesn't exist
resultsDir <- file.path(fp_out, "results")

ifelse(!dir.exists(resultsDir), 
       dir.create(resultsDir), FALSE)

# write to file errors per sample
write_csv(results_out, paste0(resultsDir, "/sample_error_results.csv"))

# write to file summary errors for whole of data run
write_csv(summaries, paste0(resultsDir, "/summary_error_results.csv"))

# write to file numerical version
write_csv(num_out, paste0(resultsDir, "/numerical_alleles.csv"))
```