---
title: "Visual Check And Export"
subtitle: "Part three of three"
author: "Bart Huntley"
date: "03/12/2019"
output: html_document
---

### Introduction
This rmarkdown document is the final of 3 that will process bat scat mass array data. The rmarkdown documents are as follows:

1. Error calculations (batscat1.Rmd)
2. Calculating groups and majority cases (batscat2.Rmd)
3. Final visualisation and export to structure format (batscat3.Rmd) - *this document*

These documents will show all code to allow for transparency and adjustment. There will be sections that require input from the **USER** prior to running. It is intended that the html output accompany any csv output to act as a reproducible record.

#### Libraries
```{r, libraries, eval=TRUE, message=FALSE, warning=FALSE}
# required packages
list.of.packages <- c("tidyverse", "kableExtra", "formattable")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# load packages
library(tidyverse)
library(kableExtra)
library(formattable)
```

#### 1 - Read in edited group majority dataset and visualise
**USER** The numerical data required for this code block was output from `batscat2.Rmd`. If you have not run that rmarkdown document please do so prior to this.

The starting point for this analysis is the `group_majorities_and_ties.csv`. This dataset may have been edited when quality assessing tied majority cases.

**USER** Please change the filepath to the `/results` directory where the data was saved to. Also change the csv name if edited version is different.
```{r, data_in, eval=TRUE, message=FALSE, warning=FALSE}
# change if required
fp_in <- "./results"
n <- "group_majorities_and_ties.csv"

# construct full name
csvdat <- file.path(fp_in, n)

# read in data and grab unique group ids
d <- read_csv(csvdat)
gps <- unique(d$group)

```

Conditional formatting is a handy visualisation technique that assists identifying possible problem data. It is uesd here to highlight any samples that have different values to the majority case for the group.

Due to the format of the data it requires a loop to cycle through the groups and multiple data formatting (long to wide, wide to long) to achieve the intended result. The majority of the conditional formatting is achieved by a custom function. This function should be readily adaptable if required.
```{r, cond_table, eval=TRUE, message=FALSE, warning=FALSE}
# empty data frame for results
tbl_df <- data.frame(stringsAsFactors = FALSE)

# function for bulk of conditional formatting
diff_to_maj <- function(x){
  ifelse(x == d1$majority,
         cell_spec(x, "html", color = "green", italic = T),
         cell_spec(x, "html", color = "red", bold = T))
}


# loop to conditionally format each group
for(i in seq_along(gps)){
  d1 <- d %>%
    dplyr::filter(group == gps[i]) %>%
    dplyr::filter(sample != "tie") %>%
    dplyr::select(-avg_amp_rate) %>%
    tidyr::pivot_longer(cols = starts_with("x"), names_to = "vars", 
                 values_to = "vals") %>%
    tidyr::pivot_wider(names_from = sample, values_from = vals) %>%
    dplyr::mutate_each(funs(replace(., is.na(.), 0)))
  
  d2 <- d1 %>%
    dplyr::mutate_at(vars(starts_with("19")), diff_to_maj) %>%
    dplyr::mutate(majority = as.character(majority)) %>%
    dplyr::mutate(majority = cell_spec(majority, background = "darkgrey", 
                                color = "white", align = "center")) %>%
    tidyr::pivot_longer(cols = c(-group, -vars), names_to = "sample", 
                 values_to = "vals") %>%
    tidyr::pivot_wider(names_from = vars, values_from = vals)
  
  tbl_df <- bind_rows(tbl_df, d2)
}

# make and view the table
tbl_df %>%
  kable(format = "html", escape = F) %>%
  kable_styling("striped", fixed_thead = T, full_width = F) %>%
  scroll_box(width = "100%", height = "800px")
```

#### 2 - Export to structure format
The last code block below will export the group majorities in *structure* format.

**USER** the csv will be saved to the `/results` directory. Change this if desired.
``````{r, export, eval=TRUE, message=FALSE, warning=FALSE}
# grab only majorities, reformat and rearrange
structure_format <- d %>%
  dplyr::filter(sample == "majority") %>%
  dplyr::mutate(g = paste0("Group", group)) %>%
  dplyr::select(g, contains("_a"), contains("_b"))

# convert NA to -9
structure_format[is.na(structure_format)] <- -9

# get rid of firts column name to match structure format
colnames(structure_format)[1] <- ""

# save to csv
fp <- "./results"
structure_out <- file.path(fp, "group_majorities_structure_format.csv")

write_csv(structure_format, structure_out)
```